The Iterative Signature Algorithm

### 2009-04-30 isa/inst/doc/ISA_tutorial.Rnw
More isa vignette improvements, use bitmaps for big image plots.

\Rfunction{isa.in.silico} returns a list with a list with three
elements, we are most interested in the first now, that is the actual
artificial data matrix. The second and the entries in the list give
the correct module memberships of the elements, see the manual page of
\Rfunction{isa.in.silico} for the details. Let's take a look at the
input data, see Fig.~\ref{fig:block}.
<<blockplot,include=FALSE,results=hide,cache=TRUE>>=
print(image(Matrix(data)))
@

\begin{figure}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<blockplot2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE,width=8,height=6>>=
<<blockplot>>
@
\caption{Another artificial data matrix with two modules and a lot of
  noisy elements.}
\label{fig:block}
\end{figure}

@@

Let us create some in-silico data to be analyzed:
<<insilico10>>=
data <- isa.in.silico(500, 200, 10)
@ 
This will be a $500\times 200$ matrix, with 10 modules. By default
``half'' of the matrix is filled with noise, so each module will be
$25\times 10$ submatrix. See the input matrix on
Fig.~\ref{fig:tenmodules}.

<<tenmods,include=FALSE,results=hide,cache=TRUE>>=
print(image(Matrix(data[[1]])))
@

\begin{figure}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<tenmods2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
<<tenmods>>
@
\caption{In-silico data with ten non-overlapping modules.}
\label{fig:tenmodules}
\end{figure}

@@

\begin{figure}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<robbox2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
<<robbox>>
@
\caption{}
\label{fig:robbox}
\end{figure}

@@

\begin{figure}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<robbox4,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
<<robbox3>>
@
\caption{}
\label{fig:robbox2}
\end{figure}

The scatterplot shows that ten modules have in fact higher
robustness scores than the rest. Let us check that these are indeed the
ten ``real'' modules of the data.
<<bestmods>>=
bestmods <- which(modules3$seeddata$rob > 61)
mod.cor <- pmin(cor(modules3$rows[,bestmods]!=0, data[[2]]),
                cor(modules3$columns[,bestmods]!=0, data[[3]]))
apply(mod.cor, 1, max)
apply(mod.cor, 2, max)
@

@@

Finally, let us take a look at some of the other modules, that were
also kept after the robustness-based filtering. Here we plot nine of
them.
<<othermods,include=FALSE,results=hide,cache=TRUE>>=
othermods <- which(modules3$seeddata$rob < 61)[1:3]
layout(matrix(seq(length(othermods)+1), 2, byrow=TRUE))
idx <- seq(ncol(data[[1]]), 1)
par(mar=c(1,1,4,1))
image(1-data[[1]][,idx], main="Input matrix", 
      axes=FALSE, frame=TRUE)
for (i in othermods) { 
  par(mar=c(1,1,4,1))
  image(1-outer(modules3$rows[,i], modules3$columns[,i])[,idx],
        main=paste("Module", i), axes=FALSE, frame=TRUE)
}
@

\begin{figure}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<othermods2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
<<othermods>>
@ 
\end{figure}

@@

<<noiseplot1,echo=FALSE,results=hide>>=
png("noise.png", width=800, height=320)
@ 
<<noiseplot2>>=
layout( matrix(1:10, nr=2, byrow=TRUE) )
for (i in seq_along(data)) {
  par(mar=c(1,1,1,1))
  image(1-data[[i]][[1]], axes=FALSE, frame=TRUE)
}
@ 
<<noiseplot3,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{noise}
  \caption{}
  \label{fig:noise}
\end{figure}

@@

<<sensbox,include=FALSE,results=hide>>=
boxplot(best, names=noise, xlab="Noise level", ylab="Sensitivity",
        cex.lab=1.5)
@
\begin{figure}
\centering
<<sensbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
<<sensbox>>
@ 
\caption{}
\label{fig:sensbox}
\end{figure}

To test the specificity, we create boxplots for the correlation
coefficients of all modules found by ISA and their closest real
module.
<<spec>>=
spec <- lapply(seq_along(modules), function(i) {
  cc <- pmin(cor(modules[[i]]$rows != 0, data[[i]][[2]]),
             cor(modules[[i]]$columns != 0, data[[i]][[3]]))
  apply(cc, 1, max)
})
@ 
<<specbox,include=FALSE,results=hide>>=
boxplot(spec, names=noise, xlab="Noise level", ylab="Specificity",
        cex.lab=1.5)
@
\begin{figure}
\centering
<<specbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
<<specbox>>
@
\caption{}
\label{fig:specbox}
\end{figure}

@@

<<overlapplot,include=FALSE,results=hide,cache=TRUE>>=
layout( rbind(1:4, c(8,5,6,7)) )
par(mar=c(1,1,3,1))
image(2-two.over[[1]], axes=FALSE, frame=TRUE, main="Input matrix")
for (i in 1:3) {
  par(mar=c(1,1,3,1))
  image(1-outer(ov.modules.1$rows[,i]!=0, ov.modules.1$columns[,i]!=0),
        axes=FALSE, frame=TRUE, main=paste("Module", i))
}
for (i in 1:3) { 
  par(mar=c(1,1,3,1))
  image( 1-outer(ov.modules.2$rows[,i]!=0, ov.modules.2$columns[,i]!=0),
        axes=FALSE, frame=TRUE, main=paste("Module", i+3))
}
@

\begin{figure}
\centering
<<overlapplot2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE,width=8,height=5>>=
<<overlapplot>>
@ 
\caption{}
\label{fig:overlapplot}
\end{figure}

>>>>

\Rfunction{isa.in.silico} returns a list with a list with three
elements, we are most interested in the first now, that is the actual
artificial data matrix. The second and the entries in the list give
the correct module memberships of the elements, see the manual page of
\Rfunction{isa.in.silico} for the details. Let's take a look at the
input data, see Fig.~\ref{fig:block}.

<<blockplot.pre,echo=FALSE,results=hide>>=
png("blockplot.png", 800, 600)
@ 
<<blockplot>>=
print(image(Matrix(data)))
@ 
<<blockplot.post,echo=FALSE,results=hide>>=
dev.off()
@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{blockplot}
\caption{Another artificial data matrix with two modules and a lot of
  noisy elements.}
\label{fig:block}
\end{figure}

@@

Let us create some in-silico data to be analyzed:
<<insilico10>>=
data <- isa.in.silico(200, 100, 10)
@ 
This will be a $\Sexpr{nrow(data[[1]])}\times \Sexpr{ncol(data[[1]])}$
matrix, with \Sexpr{ncol(data[[2]])} modules. By default ``half'' of
the matrix is filled with noise, so each module will be a 
$\Sexpr{round(nrow(data[[1]])/2/10)}\times 
\Sexpr{round(ncol(data[[1]])/2/10)}$ submatrix. See the
input matrix on Fig.~\ref{fig:tenmodules}.

<<tenmods.pre,echo=FALSE,results=hide>>=
png("tenmods.png", 600, 800)
@ 
<<tenmods>>=
print(image(Matrix(data[[1]])))
@ 
<<tenmods.post,echo=FALSE,results=hide>>=
dev.off()
@

\begin{figure}
\centering
\includegraphics{tenmods}
\caption{In-silico data with ten non-overlapping modules.}
\label{fig:tenmodules}
\end{figure}

@@

\begin{figure}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<robbox2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
<<robbox>>
@
\caption{Box-plot for the robustness scores of the modules found by
  the ISA. Some modules have a much higher score than the rest.}
\label{fig:robbox}
\end{figure}

@@

\begin{figure}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<robbox4,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
<<robbox3>>
@
\caption{The robustness scores of ISA modules after filtering out some
  modules with low robustness. It is still obvious that ten modules
  have a significantly higher score than the rest.}
\label{fig:robbox2}
\end{figure}

The scatterplot shows that ten modules have in fact higher
robustness scores than the rest. Let us check that these are indeed the
ten ``real'' modules of the data.
<<bestmods>>=
bestmods <- which(modules3$seeddata$rob > 27.5)
mod.cor <- pmin(cor(modules3$rows[,bestmods]!=0, data[[2]]),
                cor(modules3$columns[,bestmods]!=0, data[[3]]))
apply(mod.cor, 1, max)
apply(mod.cor, 2, max)
@

@@

Finally, let us take a look at some of the other modules, that were
also kept after the robustness-based filtering. Here we plot nine of
them.
<<othermods.pre>>=
png("othermods.png", 800, 800)
@ 
<<othermods>>=
othermods <- which(modules3$seeddata$rob < 27.5)[1:5]
print(plot.modules(modules3, othermods, data=data[[1]]))
@ 
<<othermods.post>>=
dev.off()
@

\begin{figure}
\centering
\includegraphics{othermods}
\caption{The input matrix and five modules with relatively low
  robustness scores. As it turns out, each of these modules contains
  in fact two blocks of the input matrix.}
\label{fig:othermods}
\end{figure}

@@

<<noiseplot.pre,echo=FALSE,results=hide>>=
png("noise.png", width=800, height=600)
@ 
<<noiseplot>>=
images(lapply(data, "[[", 1), names=as.character(noise), 
       xlab="", ylab="")
@ 
<<noiseplot.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{noise}
  \caption{Modular data sets with different levels of noise. The data
  sets are generated by adding normally distributed background noise
  with given variance to the checkerboard matrix.}
  \label{fig:noise}
\end{figure}

@@

<<sensbox,include=FALSE,results=hide>>=
boxplot(best, names=noise, xlab="Noise level", ylab="Sensitivity",
        cex.lab=1.5)
@
\begin{figure}
\centering
<<sensbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
<<sensbox>>
@ 
\caption{The sensitivity of the ISA in the function of noise. The
  sensitivity score is calculated by finding the best (in terms of
  Pearson correlation) ISA module for each block in the input matrix.

  ISA is very sensitive to relatively large noise levels.
}
\label{fig:sensbox}
\end{figure}

To test the specificity, we create boxplots for the correlation
coefficients of all modules found by ISA and their closest real
module.
<<spec>>=
spec <- lapply(seq_along(modules), function(i) {
  cc <- pmin(cor(modules[[i]]$rows != 0, data[[i]][[2]]),
             cor(modules[[i]]$columns != 0, data[[i]][[3]]))
  apply(cc, 1, max)
})
@ 
<<specbox,include=FALSE,results=hide>>=
boxplot(spec, names=noise, xlab="Noise level", ylab="Specificity",
        cex.lab=1.5)
@
\begin{figure}
\centering
<<specbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
<<specbox>>
@
\caption{Specificity of the ISA modules, in the function of
  noise. Specificity was calculated by finding the best -- again,
  by highest Pearson correlation -- matching block in the input matrix
  for each ISA module. These correlation coefficients are plotted in a
  boxplot for each noise level.

  Specificity for ISA decreases approximately linearly with the
  increase of noise.
}
\label{fig:specbox}
\end{figure}

@@

<<overlapplot.pre,echo=FALSE,results=hide>>=
png("overlapplot.png", 800, 600)
@ 
<<overlapplot>>=
plot.modules(data=two.over[[1]],
             list(rows=cbind(ov.modules.1$rows, ov.modules.2$rows),
                  columns=cbind(ov.modules.1$columns,
                    ov.modules.2$column)),
             names=c("Input matrix", paste("Module", 1:6)))
@
<<overlapplot.post,echo=FALSE,results=hide>>=
dev.off()
@

\begin{figure}
\centering
\includegraphics{overlapplot}
\caption{The input matrix with two overlapping blocks and the ISA
  modules for this data set. The first three modules were found at
  milder threshold parameters, that is why they are bigger. As you can
  see for Modules #1 and #2, at milder thresholds there is a higher
  probability that ISA will pick up some incorrect rows/columns.
}
\label{fig:overlapplot}
\end{figure}
